
# Economy

- [Income](#economy-income)

- [Poverty](#economy-poverty)

- [Employment](#economy-employment)

```{r}
#| output: false

# get acs data ----
acs_data <- readRDS(fs::path(ddir, "acs_data.rds"))
count(acs_data, variable, concept, label)

```

## Economy: Income

```{r}
#| output: false

# set graph constants ----
barvar <- "B19013_001"
titlebase <- "Median household income"
ylab <- "Dollars"
ybreaks <- seq(0, 100e3, 5e3)
data <- acs_data

```


### Washington County has lower household income than most area counties

```{r}
#| label: plot_income_county

barline2(gtype = "county", barvar, data,
         constants = constants,
         titlebase=titlebase, ylab=ylab, ybreaks=ybreaks)

```

### Cambridge village household income is about middle for nearby places

```{r}
#| label: plot_income_place

barline2(gtype = "place", barvar, data,
         constants = constants,
         titlebase=titlebase, ylab=ylab, ybreaks=ybreaks)

```


### Cambridge town household income is higher than nearby towns

```{r}
#| label: plot_income_cousub

barline2(gtype = "cousub", barvar, data,
         constants = constants,
         titlebase=titlebase, ylab=ylab, ybreaks=ybreaks)

```


## Economy: Poverty

```{r}
#| output: false

# data preparation ----

# B17001_001 total
# B17001_002 below poverty level

pov_data <- acs_data |> 
  filter(variable %in% c("B17001_001", "B17001_002")) |>
  select(stabbr, stname, geoid, geotype, geoname, trimname, variable, estimate) |> 
  pivot_wider(names_from=variable, values_from=estimate) |> 
  mutate(variable = "povrate",
         estimate = B17001_002 / B17001_001 * 100)
pov_data

```


```{r}
#| output: false

# set graph constants ----
barvar <- "povrate"
titlebase <- "Poverty rate"
ylab <- "% of population"
ybreaks <- seq(0, 100, 1)
data <- pov_data

```


### Washington County poverty rate is lower than in other area counties

```{r}
#| label: plot_povrate_county

barline2(gtype = "county", barvar,
         data,
         constants = constants,
         titlebase=titlebase, ylab=ylab, ybreaks=ybreaks)

```

### Cambridge village poverty rate is higher than in other area places

```{r}
#| label: plot_povrate_place

barline2(gtype = "place", barvar,
         data,
         constants = constants,
         titlebase=titlebase, ylab=ylab, ybreaks=ybreaks)

```


### Cambridge town poverty rate is similar to other area towns

```{r}
#| label: plot_povrate_cousub

barline2(gtype = "cousub", barvar,
         data,
         constants = constants,
         titlebase=titlebase, ylab=ylab, ybreaks=ybreaks)

```

