
```{r}
#| output: false
#| eval: true

source(here::here("slides", "setup.R"))

# get data ----
# acs_data <- readRDS(fs::path(ddir, "acs_data.rds"))
pop <- readRDS(fs::path(ddir, "population", "county_population.rds"))
placepop <- readRDS(fs::path(popdir, "popests_raw.rds"))

```


## Demographics: Population trends

### County fortunes have differed widely over last half-century

```{r}
#| label: population-levels

counties <- c("36083", "36091", "36113", "36115")

capt <-  "\nSource: U.S. Census Bureau and NYS Department of Labor"

pop |> 
  filter(geoid %in% counties) |>
  mutate(ipop=pop / pop[year==2015],
         cpcpop = ipop * 100 - 100,
         .by=geoid) |>
  ggplot(aes(x=year, y=pop / 1000)) +
  geom_line(colour = "blue", linewidth = 1) +
  scale_x_continuous(name = NULL, breaks = seq(1900, 2025, 5)) +
  scale_y_continuous(name = "Thousands of people") +
  ggtitle(label="Population in area counties") +
  labs(caption = capt) +
  theme_minimal() +
  legend_notitle +
  caption_left +
  facet_wrap(~trimname, scales="free_y")

```


### Population declines in Washington County have been persistent and severe

```{r}
#| label: population-changes

counties <- c("36083", "36091", "36113", "36115")

capt1 <-  "Source: U.S. Census Bureau and NYS Department of Labor"
capt2 <- "Note: Bump in 2020 reflects 2020 decennal census break from prior estimates"
capt <- paste0("\n", capt1, "\n", capt2)

baseyear <- 2010
pop |> 
  filter(geoid %in% counties) |>
  mutate(ipop=pop / pop[year==baseyear],
         cpcpop = ipop * 100 - 100,
         .by=geoid) |>
  filter(year >= baseyear) |>
  ggplot(aes(x=year, y=cpcpop, colour=trimname)) +
  geom_line(linewidth = 1) +
  geom_hline(yintercept = 0, linetype="solid", linewidth = .5) +
  scale_x_continuous(name = NULL, breaks = seq(baseyear, 2025, 1)) +
  scale_y_continuous(name = "Cumulative percentage change",
                     breaks =  seq(-20, 20, 1),
                     labels=scales::comma_format()) +
  ggtitle(label=paste0("Cumulative percentage change in population since ", baseyear)) +
  labs(caption = capt) +
  theme_minimal() +
  legend_notitle +
  caption_left

```


### Population declines in most area villages have been sharp

```{r}
#| output: false
#| eval: true

# add Salem CDP 
baseyear <- 2010

suppressWarnings( # Salem has fewer observations than others, generating warning
  pdata <- placepop |> 
  rename(trimname=name, pop=value) |> 
  filter(state=="36", sumlev=="157") |>  
  mutate(geoid = paste0(state, place)) |> 
  filter(geoid %in% constants$keep_places) |>
  mutate(trimname=ifelse(place=="64771", paste0(trimname, "/CDP"), trimname)) |> 
  add_row(geoid="3664771", state="36", stname="New York", county="115",
          place="64771", trimname="Salem village/CDP", year=2020, src="2020 census", pop=811) |>
  mutate(ipop=pop / pop[year==baseyear],
         cpcpop = ipop * 100 - 100,
         .by=geoid) |>
  arrange(geoid, year)
)
  

```

```{r}
#| label: village-population
#| output: true

capt1 <-  "Source: U.S. Census Bureau. Annual population estimates may not be reliable for small areas such as villages."
capt2 <- "Note: Annual estimates for Salem not easily available after 2010 due to dissolution; 2020 value is decennial census."
capt <- paste0("\n", capt1, "\n", capt2)

pdata |> 
  ggplot(aes(x=year, y=pop)) +
  geom_line(data = pdata  |>  
              filter(!(geoid=="3664771" & year > 2010)),
            colour="blue",
            linewidth = 1,
            linetype="solid") +
  geom_line(data = pdata |> 
              filter(geoid=="3664771", year >= 2010), 
            colour="blue",
            linewidth = 1,
            linetype = "dashed") + # dashed line for Salem
  scale_x_continuous(name = NULL, breaks = seq(1900, 2025, 5)) +
  scale_y_continuous(name = "Number of people") +
  ggtitle(label="Population in area villages") +
  labs(caption = capt) +
  theme_minimal() +
  legend_notitle +
  caption_left +
  facet_wrap(~trimname, scales="free_y", ncol=3)

```


:::{.notes}
Unfortunately the Census Bureau has removed annual population data after 2010 from their website as a result of the village dissolution, which occurred in 2016.
:::

### Cambridge village's population loss has not been as severe as most other area villages

```{r}
#| label: village-population-changes

capt1 <-  "Source: U.S. Census Bureau. Annual population estimates may not be reliable for small areas such as villages."
capt2 <- "Note: Annual estimates for Salem not easily available after 2010 due to dissolution; 2020 value is decennial census."
capt <- paste0("\n", capt1, "\n", capt2)

pdata2 <- pdata |> 
  filter(year >= baseyear)

pdata2 |>
  ggplot(aes(x=year, y=cpcpop, colour=trimname)) +
  geom_line(data = pdata2  |>  
              filter(!(geoid=="3664771")),
            linewidth = 1,
            linetype="solid") +
  geom_line(data = pdata2 |> 
              filter(geoid=="3664771"), 
            linewidth = 1,
            linetype = "dashed") + # dashed line for Salem  
  geom_hline(yintercept = 0, linetype="solid", linewidth = .5) +
  scale_x_continuous(name = NULL, breaks = seq(baseyear, 2025, 1)) +
  scale_y_continuous(name = "Cumulative percentage change",
                     breaks =  seq(-20, 20, 1),
                     labels=scales::comma_format()) +
  ggtitle(label=paste0("Cumulative percentage change in population since ", baseyear)) +
  labs(caption = capt) +
  theme_minimal() +
  legend_notitle +
  caption_left

```


### Child poverty

```{r}
# E:\R_projects\projects\cambridge\data\saipe
# ussd23.txt
#           1- 2       FIPS State code (00 for US record)
        #   4- 8       District ID
        #  10-81       District Name
        #  83-90       Total Population 
        #  92-99       Population of Relevant Children 5 to 17 years of Age
        # 101-108      Estimated Number of Relevant Children 5 to 17 years old
        #              in Poverty Related to the Householder 
        # 110-130      A tag indicating the file name and date of creation

starts <- c(1, 4, 10, 83, 92, 101, 110)
ends <- c(2, 8, 81, 90, 99, 108, 130)
cnames <- c("stfips", "distid", "distname", "totpop", "childpop", "childpoverty", "tag")
fpath <- fs::path(ddir, "saipe", "ussd23.txt")

df <- vroom_fwf(fpath, 
                col_positions = fwf_positions(start=starts, end=ends, 
                col_names=cnames)) |> 
  select(-c(tag)) |> 
  mutate(povrate = childpoverty / childpop * 100)

check <- df |> 
  filter(stfips == "36")

# argyle 03210
# cambridge 06210
# greenwich 12900
# salem 25470
# schuylerville 26160
# hoosick falls 14760
 
keep_sds <- c("03210", "06210", "12900", "25470", "26160", "14760")

pdata <- df |> 
  filter(stfips == "36", distid %in% keep_sds) |> 
  mutate(trimname=case_when(distid=="06210" ~ "Cambridge",
                            distid=="03210" ~ "Argyle",
                            distid=="12900" ~ "Greenwich",
                            distid=="25470" ~ "Salem",
                            distid=="26160" ~ "Schuylerville",
                            distid=="14760" ~ "Hoosick Falls")) |> 
  arrange(distid)


```

